White Wine Quality by Kutay Erkan
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(tidyverse)
library(rmarkdown)
library(knitr)
library(reshape2)
library(psych)
library(scales)
library(RColorBrewer)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
raw = read.csv("wineQualityWhites.csv")
```

In this report, we will try to explore a dataset with approximately 4900 
observations about the attributes of white wine and its perceived quality.

# Univariate Plots Section

In our dataset, we have 4898 observations and 13 columns. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Summary}
str(raw)
summary(raw)
```

Our initial observations are:

* There are 4898 observations and 13 variables.
* Quality and X are integer variables.
* Quality is the variable we will focus on the most. It is an integer (discrete) 
and categorical variable with values between 3 and 9. Its mean is 5.878 and 
median is 6.
* X seems to be a numerical Unique Identifier Variable, as difference between 
consecutive variables is always 1.


```{r message=FALSE, warning=FALSE, Check_UID}
all(diff(raw$x) == 1)
```
we also check to see if there are any non-positive values in our dataframe: 

```{r message=FALSE, warning=FALSE, Check_Positivity}
colSums(raw == 0)
colSums(raw < 0)
```
We see that only citric.acid contains 19 zero values. It is curious if this is 
because it is not reported or those observations really contained zero values, 
and we will keep this in mind.

We will draw quick histograms for all variables to get a general idea of the 
variables, using X as the unique id. We'll free the x axis so that histograms 
of different variables become meaningful.

```{r echo=FALSE, message=FALSE, warning=FALSE, class_remover}
#raw = subset(raw, select = -class -acidity)
#used if it is needed to remove the added features
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
melt_raw = melt(raw, id.vars = "X")
ggplot(melt_raw, aes(x = value)) + 
    facet_wrap(~variable, scales = "free_x") + 
    geom_histogram()
```

From now on we don't need X as a unique id in the melted histograms above, so 
we can set it to NULL.

```{r message=FALSE, warning=FALSE, Sex_X_Null}
raw$X = NULL
```

We also see that as quality is a discrete variable, it can be better shown is the 
following way with geom_bar (we once again see that minimum quality is 3 and 
maximum is 9):

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Plot}
ggplot(raw, aes(x = quality)) + 
  geom_bar() +
  scale_x_discrete(limits=0:10)
```

# Univariate Analysis
### Transformations
When we look at our histograms, we see that following variables seem skewed:
* fixed.acidity
* volatile.acidity
* citric.acid
* free.sulfur.dioxide
* total.sulfur.dioxide
* sulphates
* alcohol

When we research it, we learn that pH scale is approximately base 10 logarithm 
of hydrogen ions in a solution. Considering this fact and the tails, we will
transform *.acidity variables using base 10 logarithm.
```{r echo=FALSE, message=FALSE, warning=FALSE, Log_Transforms}
ggplot(data = raw,
       aes(x = fixed.acidity)) +
  geom_histogram() +
  scale_x_log10()

ggplot(data = raw,
       aes(x = volatile.acidity)) +
  geom_histogram() +
  scale_x_log10()

ggplot(data = raw,
       aes(x = citric.acid)) +
  geom_histogram() +
  scale_x_log10()
```

We get very normal distributions except for the tail of citric.acid, containing 
values lower than 0.1 g/dm^3.

### Outliers

We use boxplots with an alpha value of 0.05 to see possible outliers in each of 
our datasets.

```{r echo=FALSE, message=FALSE, warning=FALSE, Boxplot_Outliers}
ggplot(data = melt_raw, aes(x = variable, y = value)) +
  geom_boxplot(aes(fill = X), outlier.alpha = 0.05) +
  facet_wrap( ~ variable, scales = "free")
```

### New (Created) Features
As quality variable is basically a measure of how good or bad a wine is, it is 
logical to convert it to a categorical variable. For this purpose we create a 
new variable called "class". We can use class to separately inspect what is 
happening feature-wise in different types of wine (bad, good etc.).

While separating quality levels to create a "class" variable, at first I thought
using 3 classes; "bad", "good", and "average". But I decided against it as 
bundling quality levels 5 and 6 together would cause a huge fraction of data 
points to be in the same class. Thus it is decided to use four levels: A, B, C 
and D. Grouping can be seen below:

|   Quality   | Class |
|:-----------:|:-----:|
| 7, 8, 9, 10 |   A   |
|      6      |   B   |
|      5      |   C   |
|  1, 2, 3, 4 |   D   |

```{r echo=FALSE, message=FALSE, warning=FALSE, Categorize_Quality}
raw$class = cut(raw$quality, 
                          c(0.0,4.5,5.5,6.5,10.0),
                          labels=c('D','C','B', 'A'))
```

We already have a variable for pH, but when we already have 3 variables for 
acidity, we can assign a separate variable for acidity and use it as a feature.

Looking in the documentation, unit for all acidity variables are g/dm^3, so we 
can directly sum them.

```{r echo=FALSE, message=FALSE, warning=FALSE, Create_Acidity}
raw$acidity = raw$fixed.acidity + raw$volatile.acidity + raw$citric.acid
```

# Bivariate Plots Section

To get a quick snapshot of relationships between all the fatures, we look at 
their plot and correlation matrices:

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
pairs.panels(subset(raw, select = -class))
```


When we look at the relationship between quality and other features, there is a 
strong relationship between quality and alcohol. Density is correlated with 
alcohol (inversely) and residual sugar, which is not surprising. We also see
that quality is correlated with chlorides (salt), volatile acidity, and total 
sulfur dioxide; all of which will be examined further.

When we use jitter and transparency on the points we see alcohol has a 
significant positive effect on the quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_2}
ggplot(aes(x = alcohol, y = quality), data = raw) +
  geom_jitter(alpha = 0.3, width = 0.3, color="orange")
```


While chlorides don't seem to "break" a wine in lower amounts, the fact that 
there is no wine with a quality of 6 or bigger when chloride (salt) amount is 
bigger than 0.2 g/dm^3 might indicate higher amounts of salt affects quality 
perception. Indeed, top notch wines (quality of 9) have lower than 0.05 g/dm^3.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_3}
ggplot(aes(x = chlorides, y = quality), data = raw) +
  geom_jitter(height = 0.1, color="orange")
```

Effect of volatile acidity on quality is not exactly clear, and should be 
further investigated later on.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_4}
ggplot(aes(x = volatile.acidity, y = quality), data = raw) +
  geom_jitter(height = 0.1, color="orange") +
  scale_x_log10()
#  scale_x_continuous(lim = 0, quantile(raw$volatile.acidity, 0.95)))
```

Even though we remove outliers of bottom and top by 5%, there's no significant 
change on quality created by different sulfur dioxide amounts.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_5}
ggplot(aes(x = total.sulfur.dioxide, y = quality), data = raw) +
  geom_jitter(height = 0.1, color="orange") +
  scale_x_continuous(lim = c(quantile(raw$total.sulfur.dioxide, 0.05),
                             quantile(raw$total.sulfur.dioxide, 0.95)))
```

# Bivariate Analysis

Other than obvious relationships between the ones like sulfur.dioxides or 
acidities; strongest relationships are between density - alcohol, 
density - residual sugar, and quality - alcohol. It is also interesting that 
fixed acidity has a 0.99 correlation with acidity, suggesting fixed acids are 
in much higher concentrations in the wines.

# Multivariate Plots Section

we use the classes we created to differentiate points, and apply upper 95% 
quantile on chlorides.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
ggplot(data = raw, aes(x = alcohol, y = chlorides,
           color = class)) +
  geom_point() +
  scale_color_brewer(type = 'seq', palette = 5,
    guide = guide_legend(title = 'Class', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  scale_y_continuous(limits = c(0, quantile(raw$chlorides, 0.95)))
```

We see in the documentation that "volatile acidity" is acetic acid. If a wine has
both low alcohol and high acetic acid, it tends to have worse quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_2}
ggplot(data = raw, aes(x = alcohol, y = volatile.acidity,
           color = class)) +
  geom_point() +
  scale_color_brewer(type = 'seq', palette = 5,
    guide = guide_legend(title = 'Class', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  scale_y_continuous(trans = log10_trans())
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_3}
ggplot(data = raw, aes(x = alcohol, y = total.sulfur.dioxide,
           color = class)) +
  geom_point(alpha = 0.5) +
  scale_color_brewer(type = 'seq', palette = 5,
    guide = guide_legend(title = 'Class', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  scale_y_continuous(limits = c(0, quantile(raw$total.sulfur.dioxide, 0.95)))
```

# Multivariate Analysis

We see that alcohol is indeed very important in the perceived quality of a wine.
Having a high salt concentration affects it badly, and other chemicals such as 
volatile acids and sulfur dioxide shouldn't be higher than a certain amount.


### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(aes(x=alcohol, y=quality),data=raw) +
  geom_jitter(alpha = 0.3, width = 0.4, color="orange") +
  geom_smooth(method = lm) +
  labs(x = "Alcohol (% by Volume)", y = "Quality") +
  ggtitle('White Wine Quality by Alcohol')
```

### Description One

This is the first clear indicator we have regarding the relationship between 
alcohol content and perceived quality of a white wine.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(data = raw, aes(x = alcohol, y = chlorides,
           color = class)) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  scale_color_brewer(type = 'seq', palette = 5,
    guide = guide_legend(title = 'Class', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  scale_y_continuous(limits = c(0, quantile(raw$chlorides, 0.95))) +
  labs(x = "Alcohol (% by Volume)",
       y = "Chlorides (sodium chloride - g / dm^3)") +
  ggtitle('White Wine Quality by Alcohol and Chloride')
```

### Description Two

We see that Class A and B wines mostly reside in the lower right corner while 
lower quality wines are generally in upper left corner. In addition to alcohol, 
we see that high amount of salt might affect a wine's quality badly.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(data = raw, aes(x = alcohol, y = volatile.acidity,
           color = class)) +
  geom_point(alpha = 1, size = 1.5, position = 'jitter') +
  scale_color_brewer(type = 'seq', palette = 5,
    guide = guide_legend(title = 'Class', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  scale_y_continuous(trans = log10_trans()) +
  labs(x = "Alcohol (% by Volume)",
       y = "log10 of Acetic Acid (g / dm^3)") +
  ggtitle('White Wine Quality by Alcohol and log10 Trans. of Acetic Acid')
```

### Description Three

Finally, we also see acetic acid affecting the quality of a wine in addition to 
alcohol. Note the log10 transformation, used as volatile.acidity variable a 
form close to normal when transformed, and it is naturally a logarithmic 
variable.

# Reflection

Main struggle for me was having an integer variable (quality) to investigate 
other features' effects on. It required more technical additions, like 
converting it to a ordinal variable. Also not having much of an intutition 
regarding which material(s) make a wine better prompted me to make extra 
research. Seeing how alcohol played a part in wine quality also surprised me. A 
point going forward could be using a mathematical model to test if the 
mentioned relationships are there, and can such models predict "quality" of a 
wine without no one testing it.